New cardinality estimation algorithms for HyperLogLog sketches

### 2016-12-05 paper/paper.tex
improved future work section, discussion of one permutation hashing, added more references

In case the data is not distributed and results do not need to be aggregated further, there are even more efficient estimation algorithms available. On the one hand there is the self-learning bitmap \cite{Chen2011}, and on the other hand there is the HyperLogLog algorithm extended by a historic inverse probability estimator that is continuously updated together with the HyperLogLog sketch \cite{Ting2014}. Both achieve the same estimation error using less space. However, the estimated cardinality depends on the insertion order of elements and hence cannot be used in a distributed environment.

@@

\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\textstyle\frac{1}{2}
\symDistanceMeasure
+
\textstyle\frac{1}{4}
\symDistanceMeasure^2
\textstyle\frac{
\symPoissonRate_\symSetASuffix
\symPoissonRate_\symSetBSuffix
}{\left(
\symPoissonRate_\symSetASuffix+
\symPoissonRate_\symSetBSuffix
\right)^2}
\right)
\end{equation}
where $\symDistanceMeasure = \frac{\symPoissonRate_\symSetASuffix
+\symPoissonRate_\symSetBSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}$
 is the Jaccard distance. Since $
\frac{
\symPoissonRate_\symSetASuffix
\symPoissonRate_\symSetBSuffix
}{\left(
\symPoissonRate_\symSetASuffix+
\symPoissonRate_\symSetBSuffix
\right)^2}
$ is always in the range $[0,\frac{1}{4}]$, the probability for equal register values can be bounded by
\begin{equation}
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\textstyle\frac{1}{2}
\symDistanceMeasure
\right)
\lesssim
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\lesssim
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\textstyle\frac{1}{2}
\symDistanceMeasure
+
\textstyle\frac{1}{16}
\symDistanceMeasure^2
\right).
\end{equation}
As shown in \cref{fig:equal_register_probability} the bounds are very close, especially for small Jaccard distances where the probability can be well approximated by
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1-
\symAlpha_\infty\symDistanceMeasure.
\end{equation}
\begin{figure}
\centering
\includesvg[width=0.6\textwidth]{equal-register-probability}
\caption{The approximate probability range of equal register values as a function of the Jaccard distance.}
\label{fig:equal_register_probability}
\end{figure}
This dependency on the Jaccard distance is very similar to that of minwise hashing \eqref{equ:minwise_hahsing_probability}, which makes the
HyperLogLog algorithm an interesting candidate for locality-sensitive hashing with respect to the Jaccard similarity. The memory-efficiency of different hashing approaches can be measured by a storage factor that is the variance of the distance estimator multiplied by the number of bits used for storing the hash signature \cite{Li2011}. For a hash algorithm that maps two different sets to the same $\symNumBitsMinwiseHashing$-bit hash value with probability $1-\symA\symDistanceMeasure$ with some constant $\symA\in(0,1]$, the storage factor is given by $\symNumBitsMinwiseHashing\symDistanceMeasure\left(\frac{1}{\symA}-\symDistanceMeasure\right)$. 
Particularly, for the case $D=0.3$, this gives \num{6.72} for conventional minwise hashing with 32-bit hash values ($\symNumBitsMinwiseHashing=32$, $\symA\approx 1$), \num{0.51} for 1-bit minwise hashing ($\symNumBitsMinwiseHashing=1$, $\symA=\frac{1}{2}$), and  \num{1.63} for the HyperLogLog algorithm with 5-bit registers ($\symNumBitsMinwiseHashing=5$, $\symA=
\symAlpha_\infty$). This means that the HyperLogLog algorithm is not as memory-efficient as 1-bit minwise hashing, but significantly better than 32-bit minwise hashing.

It is worth mentioning that estimating the Jaccard distance by just counting equal registers does not use all the information contained by the HyperLogLog sketch. The method described in \cref{sec:cardinality_estimation_set_intersections} would allow a more accurate estimation of the Jaccard distance by using the estimates for intersection and union sizes. This could be used for additional more precise filtering when searching for similar items.

Preprocessing is very costly for minwise hashing, because many hash function evaluations are necessary for all the needed permutations \cite{Li2011}. In contrast, the HyperLogLog algorithm requires only a single hash function evaluation which reduces preprocessing costs. In addition, HyperLogLog sketches can be easily merged, which is not possible for $\symNumBitsMinwiseHashing$-bit minwise hash signatures.

\section{Conclusion}

We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches based on the Poisson approximation. For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation. The first uses the original estimator extended by theoretically motivated correction terms. The second is based on the maximum likelihood method and solves the corresponding optimization problem using the secant method. Both algorithms are inherently unbiased and can be implemented without dependence on empirically determined bias correction data.

>>>>

In case the data is not distributed and results do not need to be aggregated further, there are even more efficient estimation algorithms available. On the one hand there is the self-learning bitmap \cite{Chen2011}, and on the other hand there is the HyperLogLog algorithm extended by a historic inverse probability estimator that is continuously updated together with the HyperLogLog sketch \cite{Ting2014, Cohen2014}. Both achieve the same estimation error using less space. However, the estimated cardinality depends on the insertion order of elements and hence cannot be used in a distributed environment.

@@

\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\textstyle\frac{1}{2}
\symDistanceMeasure
+
\textstyle\frac{1}{4}
\symDistanceMeasure^2
\textstyle\frac{
\symPoissonRate_\symSetASuffix
\symPoissonRate_\symSetBSuffix
}{\left(
\symPoissonRate_\symSetASuffix+
\symPoissonRate_\symSetBSuffix
\right)^2}
\right)
\end{equation}
where $\symDistanceMeasure = \frac{\symPoissonRate_\symSetASuffix
+\symPoissonRate_\symSetBSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}$
 is the Jaccard distance. Since $
\frac{
\symPoissonRate_\symSetASuffix
\symPoissonRate_\symSetBSuffix
}{\left(
\symPoissonRate_\symSetASuffix+
\symPoissonRate_\symSetBSuffix
\right)^2}
$ is always in the range $[0,\frac{1}{4}]$, the probability for equal register values can be bounded by
\begin{equation}
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\textstyle\frac{1}{2}
\symDistanceMeasure
\right)
\lesssim
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\lesssim
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\textstyle\frac{1}{2}
\symDistanceMeasure
+
\textstyle\frac{1}{16}
\symDistanceMeasure^2
\right).
\end{equation}
As shown in \cref{fig:equal_register_probability} the bounds are very close, especially for small Jaccard distances where the probability can be well approximated by
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1-
\symAlpha_\infty\symDistanceMeasure.
\end{equation}
\begin{figure}
\centering
\includesvg[width=0.6\textwidth]{equal-register-probability}
\caption{The approximate probability range of equal register values as a function of the Jaccard distance.}
\label{fig:equal_register_probability}
\end{figure}
This dependency on the Jaccard distance is very similar to that of minwise hashing \eqref{equ:minwise_hahsing_probability}, which makes the
HyperLogLog algorithm an interesting candidate for locality-sensitive hashing with respect to the Jaccard similarity. The memory-efficiency of different hashing approaches can be measured by a storage factor that is the variance of the distance estimator multiplied by the number of bits used for storing the hash signature \cite{Li2011}. For a hash algorithm that maps two different sets to the same $\symNumBitsMinwiseHashing$-bit hash value with probability $1-\symA\symDistanceMeasure$ with some constant $\symA\in(0,1]$, the storage factor is given by $\symNumBitsMinwiseHashing\symDistanceMeasure\left(\frac{1}{\symA}-\symDistanceMeasure\right)$. 
Particularly, for the case $D=0.3$, this gives \num{6.72} for conventional minwise hashing with 32-bit hash values ($\symNumBitsMinwiseHashing=32$, $\symA\approx 1$), \num{0.51} for 1-bit minwise hashing ($\symNumBitsMinwiseHashing=1$, $\symA=\frac{1}{2}$), and  \num{1.63} for the HyperLogLog algorithm with 5-bit registers ($\symNumBitsMinwiseHashing=5$, $\symA=
\symAlpha_\infty$). This means that the HyperLogLog algorithm is not as memory-efficient as 1-bit minwise hashing, but significantly better than 32-bit minwise hashing, when estimating the Jaccard distance by just counting equal registers. However, in contrast to 1-bit minwise hashing the HyperLogLog sketch contains additional information which allows estimating the cardinality or merging two HyperLogLog hash signatures. Furthermore, the method described in \cref{sec:cardinality_estimation_set_intersections} would allow a more accurate estimation of the Jaccard distance by using the estimates for intersection and union sizes. This could be used for additional more precise filtering when searching for similar items.

Since HyperLogLog sketches can be effectively constructed, because only a single hash function evaluation is needed for each item, the preprocessing step would be very fast. In contrast, preprocessing is very costly for minwise hashing, because of the many required permutations \cite{Li2011}. To overcome this problem, one permutation hashing was proposed \cite{Li2012} which keeps for a predefined number of bins the minimum of all values that are mapped to this bin. Actually, this is very similar to the HyperLogLog algorithm. The bins correspond to HyperLogLog registers which keep the first one-bit position of the minimum value instead of the minimum value itself. This close connection was also pointed out in \cite{Cohen2014}, where both methods have been considered as variants of $k$-partion minwise hashing. The only difference is that the HyperLogLog algorithm uses base-2 ranks while one permutation hashing uses full ranks.

\section{Conclusion}
We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches based on the Poisson approximation. For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation. The first uses the original estimator extended by theoretically motivated correction terms. The second is based on the maximum likelihood method and solves the corresponding optimization problem using the secant method. Both algorithms are inherently unbiased and can be implemented without dependence on empirically determined bias correction data.
